Object-oriented implementations of the MPDATA advection equation solver in~C++,~Python~and~Fortran

### 2013-03-07 paper/doc/paper.tex
some paper.tex changes

  Using an additional array dimension to represent the sequence elements is not considered for two reasons.
  First, the arrays constituting the sequence elements may have different sizes
      (e.g.  $C^{[x]}$ and $C^{[y]}$ on the Arakawa-C grid discussed in section~\ref{sec:grid}).
  Second, the order of dimensions would need to be different for different languages to assure that
      the contiguous dimension is not used for time levels.

@@

    \draw [ultra thick,-latex,red] (0,.8) -- (0,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (0,-1.2) -- (0,-.8) node [below right] {$C^{[y]}_{[i,j\mhlf]}$};
    \draw [ultra thick,-latex,red] (-2,.8) -- (-2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (-2,-1.2) -- (-2,-.8) node [above] {};
    \draw [ultra thick,-latex,red] (2,.8) -- (2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (2,-1.2) -- (2,-.8) node [above] {};
  \end{tikzpicture}
  \caption{\label{fig:grid}
    A schematic of the Arakawa-C grid.
  }
  \end{figure}
  The so-called Arakawa-C staggered grid \citep{Arakawa_and_Lamb_1977} depicted in Figure~\ref{fig:grid}.
    is a natural choice for MPDATA.
  MPDATA is a finite-difference approximation to the flux-form of the equation~(\ref{eq:adv}) 
    resulting in a finite-volume scheme.
  Using the Arakawa-C grid allows to evaluate the fluxes at 
    the grid cell (finite volume) edges without interpolation.

  As a consequence, the discretised representations of the $\psi$ scalar field, and each 
    component of the ${\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}}$ vector field 
    in eq.~(\ref{eq:adv}) are defined over different grid point locations.
  In mathematical notation this can be indicated by usage of fractional indices, e.g.
    $C^{[x]}_{[i\mhlf,j]}$, $C^{[x]}_{[i+\phlf,j]}$, $C^{[y]}_{[i,j\mhlf]}$ and $C^{[y]}_{[i,j+\phlf]}$
    to depict the grid values of the ${\vec{C}}$ vector components surrounding $\psi_{[i,j]}$.
  However, fractional indexing does not have a built-in counterpart in any of the
    employed programming languages.
  A desired syntax would translate:
  \begin{itemize}
    \item{\prog{$i-\phlf$} ~~ $\leadsto$ ~~ \prog{$i-1$}} 
    \item{\prog{$i+\phlf$} ~~ $\leadsto$ ~~ \prog{$i$}.}
  \end{itemize}
  OOP offers a convenient way to implement such notation
    by overloading the \prog{+} and \prog{-} operators for objects representing array indices.

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice}/\prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{Shift} class with the adequate operator overloads:
  \codepyt{code/pyt/listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}one could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate class}:
  \codepyt{code/pyt/listings.py}{listing04}{listing05}{1}

  In Fortran, similarly as in C++ and Python, the logic related to employment of the Arakawa-C grid
    allowing employment of fractional indices in array indexing is obtained through
    definition and instantiation of an object representing the half, and having appropriate
    operator overloads:
  \codefor{code/for/listings.f}{listing03}{listing04}{1}

@@

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t} constructed with a permutation
    of the \prog{i} and \prog{j} arguments of \prog{pi()}, with permutation order
    dependant on the template parameter \prog{d}:
  \codecpp{code/cpp/listings.hpp}{listing07}{listing08}{1}

@@

  In the Fortran implementation an analogous behaviour is obtained by introducing
    a function \prog{pi()} returning a pointer the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code/for/listings.f}{listing05}{listing06}{1}

@@

  The authors do expect some performance gain could readily
    be obtained by introducing into the codes what is referred to as
    {\em premature optimisation} -- code rearrangements aimed solely
    at the purpose of increasing performance.
  These were avoided intentionally as they degrade code readability,
    should in principle be handled by the compilers,
    and are generally advised to be avoided \citep[e.g.][section 3.12]{bib_CERNcpp}.

@@

  Python simplifies also the building and dissemination of software,
    especially if multiple architectures and operating systems are targeted.
  However, CMake is a tool that allows to efficiently automate builds, tests and 
    packaging of C++ and Fortran programs.

@@

  The performance evaluation revealed that:
    \begin{itemize}
      \item{the Fortran set-up offered shortest execution times,}
      \item{it took the C++ set-up less than twice longer to compute than Fortran,}
      \item{C++ and PyPy set-ups offered significantly smaller memory consumption 
        than Fortran and CPython,}
      \item{the PyPy set-up was roughly twice slower than C++ and up to twice faster than CPython.}
    \end{itemize}
  The three equally-structured implementations required ca. 200, 300, and 550 lines of code
    in Python, C++ and Fortran, respectively.

>>>>

  Using an additional array dimension to represent the sequence elements is not considered for two reasons.
  First, the $C^{[x]}$ and $C^{[y]}$ arrays constituting the sequence have different sizes
    (see discussion of the Arakawa-C grid in section~\ref{sec:grid}).
  Second, the order of dimensions would need to be different for different languages to assure that
      the contiguous dimension is not used for time levels.

@@

    \draw [ultra thick,-latex,red] (0,.8) -- (0,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (0,-1.2) -- (0,-.8) node [below right] {$C^{[y]}_{[i,j\mhlf]}$};
    \draw [ultra thick,-latex,red] (-2,.8) -- (-2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (-2,-1.2) -- (-2,-.8) node [above] {};
    \draw [ultra thick,-latex,red] (2,.8) -- (2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (2,-1.2) -- (2,-.8) node [above] {};
  \end{tikzpicture}
  \caption{\label{fig:grid}
    A schematic of the Arakawa-C grid.
  }
  \end{figure}
  The so-called Arakawa-C staggered grid \citep{Arakawa_and_Lamb_1977} depicted in Figure~\ref{fig:grid}
    is a natural choice for MPDATA.
%  MPDATA is a finite-difference approximation to the flux-form of the equation~(\ref{eq:adv}) 
%    resulting in a finite-volume scheme.
%  Using the Arakawa-C grid allows to evaluate the fluxes at 
%    the grid cell (finite volume) edges without interpolation.
  As a consequence, the discretised representations of the $\psi$ scalar field, and each 
    component of the ${\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}}$ vector field 
    in eq.~(\ref{eq:adv}) are defined over different grid point locations.
  In mathematical notation this can be indicated by usage of fractional indices, e.g.
    $C^{[x]}_{[i\mhlf,j]}$, $C^{[x]}_{[i+\phlf,j]}$, $C^{[y]}_{[i,j\mhlf]}$ and $C^{[y]}_{[i,j+\phlf]}$
    to depict the grid values of the ${\vec{C}}$ vector components surrounding $\psi_{[i,j]}$.
  However, fractional indexing does not have a built-in counterpart in any of the
    employed programming languages.
  A desired syntax would translate:
  \begin{itemize}
    \item{\prog{$i-\phlf$} ~~ $\leadsto$ ~~ \prog{$i-1$}} 
    \item{\prog{$i+\phlf$} ~~ $\leadsto$ ~~ \prog{$i$}.}
  \end{itemize}
  OOP offers a convenient way to implement such notation
    by overloading the \prog{+} and \prog{-} operators for objects representing array indices.

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{Shift} class with the adequate operator overloads:
  \codepyt{code/pyt/listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}one could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate class}:
  \codepyt{code/pyt/listings.py}{listing04}{listing05}{1}

  In Fortran fractional array indexing is obtained through
    definition and instantiation of an object representing the half, and having appropriate
    operator overloads:
  \codefor{code/for/listings.f}{listing03}{listing04}{1}

@@

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}:
  \codecpp{code/cpp/listings.hpp}{listing07}{listing08}{1}
  (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations).

@@

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code/for/listings.f}{listing05}{listing06}{1}

@@

  The authors do expect some performance gain could 
    be obtained by introducing into the codes some ''manual'' optimisations -- 
    code rearrangements aimed solely at the purpose of increasing performance.
  These were avoided intentionally as they degrade code readability,
    should in principle be handled by the compilers,
    and are generally advised to be avoided \citep[e.g.][section 3.12]{bib_CERNcpp}.

@@

  Python simplifies also the building and dissemination of software,
    especially if multiple architectures and operating systems are targeted.
  However, CMake\footnote{CMake is a family of open-source, cross-platform
    tools automating building, testing and packaging of C/C++/Fortran software,
    see http://cmake.org/ for details}
    is a tool (or arguably the tool) that allows to efficiently automate 
    builds, tests and packaging of C++ and Fortran programs.

@@

  The performance evaluation revealed that:
    \begin{itemize}
      \item{the Fortran set-up offered shortest execution times,}
      \item{it took the C++ set-up less than twice longer to compute than Fortran,}
      \item{C++ and PyPy set-ups offered significantly smaller memory consumption 
        than Fortran and CPython,}
      \item{the PyPy set-up was roughly twice slower than C++ and up to twice faster than CPython.}
    \end{itemize}
  The three equally-structured implementations required ca. 200, 300, and 550 lines of code %TODO update line numbers
    in Python, C++ and Fortran, respectively.
